---
title: "Covid 19 data"
author: "Ernesto Medina"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("stringr") 
library("readr")
library(dplyr)
library(tidyr)
library(tidyverse)
library(lubridate)
library(gridExtra)
```

## Covid-19, is significant the relation between cases and deaths?


To answer this question we'll be analyzing a data source with covid data from the beginning of reporting data.

The source of the data comes from the COVID-19 Data Repository by the Center for Systems Science and Engineering (CSSE) at Johns Hopkins University: <https://github.com/CSSEGISandData/COVID-19>.

We will be importing the time series covid19 data.

```{r cars}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv","time_series_covid19_deaths_US.csv","time_series_covid19_confirmed_global.csv","time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)
urls
```

### Import data

```{r import_data, message=FALSE}
cases_us <- read_csv(urls[1])
deaths_us <- read_csv(urls[2])
global_cases <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
```

## Data Description

In this document we are presenting covid 19 data from Johns Hopkins University gihub repo. We will go through a process of visual analysis and modeling. The data we import is as follows:

-   Data collected daily and updated every day.

-   The data consists of 3342,3342,285 and 285 rows in the datasets.

-   Column names are Province/State\|Country/Region\|Lat/Long\|and each day from 1/22/20.

-   We won't need Lat Lon and will need to "tidy" the date columns to our needs.We will make each date column into a row.

## Data Transformation

```{r pivot_datasets}
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province/State`,`Country/Region`,Lat,Long),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long))
global_deaths <- global_deaths %>%
  pivot_longer(cols = -c(`Province/State`,`Country/Region`,Lat,Long),
               names_to = "date",
               values_to = "deaths") %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat,Long))
cases_us <- cases_us %>%
  pivot_longer(cols = -c(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  mutate(date = mdy(date)) %>%
  select(Admin2:cases) %>%
  select(-c(Lat,Long_))
deaths_us <- deaths_us %>%
  pivot_longer(cols = -c(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  mutate(date = mdy(date)) %>%
  select(Admin2:deaths) %>%
  select(-c(Lat,Long_))
```

```{r join_datasets}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region= `Country/Region`,
         Province_State=`Province/State`)
us <- cases_us %>%
  full_join(deaths_us) 
```

```{r fill_data}
global <- global %>% 
  unite("Combined_Key",
        c(Province_State,Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
uid_lookup_url<- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat,Long_,Combined_Key, code3, iso2,iso3,Admin2))
global <- global %>%
  left_join(uid, by = c("Province_State","Country_Region")) %>%
  select(-c(UID, FIPS)) 
```

##Visualize Data


```{r prepare_spource_for_visualization}

us_by_state <- us %>%
  group_by(Province_State,Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
                        Population = sum(Population)) %>%
  mutate(deaths_per_thou = deaths *1000/Population,
         cases_per_thou = cases *1000/Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_thou, cases_per_thou, Population) %>%
  ungroup()

us_totals <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
                        Population = sum(Population)) %>%
  mutate(deaths_per_thou = deaths *1000/Population,
         cases_per_thou = cases *1000/Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_thou, cases_per_thou, Population) %>%
  ungroup()
  
global_by_country <- global %>%
  group_by(Country_Region,date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
                        Population = sum(Population)) %>%
  mutate(deaths_per_thou = deaths *1000/Population,
         cases_per_thou = cases *1000/Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_thou, cases_per_thou, Population) %>%
  ungroup()
global_total <- global %>%
  group_by(date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
                        Population = sum(Population)) %>%
  mutate(deaths_per_thou = deaths *1000/Population,
         cases_per_thou = cases *1000/Population) %>%
  select(date, cases, deaths, deaths_per_thou, cases_per_thou, Population) %>%
  ungroup()
us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(cases = max(cases), deaths = max(deaths),
                        Population = max(Population),
            deaths_per_thou = deaths *1000/Population,
         cases_per_thou = cases *1000/Population) %>%
  filter(cases > 0, Population > 0)
```


```{r visualization_us_totals}
us_totals %>%
  ggplot(aes(x = date, y= cases)) +
  xlab("Date") +
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths,color="deaths"))+
  geom_point(aes(y=deaths,color="deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "Covid 19 in US", y=NULL)
```



```{r visualization_global}
global_total %>%
  ggplot(aes(x = date, y= cases)) +
  xlab("Date") +
  geom_line(aes(color="cases"))+
  geom_point(aes(color="cases")) +
  geom_line(aes(y=deaths,color="deaths"))+
  geom_point(aes(y=deaths,color="deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "Covid 19 global", y=NULL)
```
## Analyzing data
```{r ranks}
rank_us <- us_state_totals %>%
  mutate(rank = rank(-cases), rank_deaths = rank(-deaths), rank_mill = rank(-cases_per_thou),
         rank_deaths_mill = rank(-deaths_per_thou))
rank_us <- rank_us[order(rank_us$rank),]
rank_by_death_us <- rank_us[order(rank_us$rank_deaths),]
rank_us_thousand <- rank_us[order(rank_us$rank_mill),]
rank_by_death_us_thousand <- rank_us[order(rank_us$rank_deaths_mill),]
```
The US State with more total cases: `r rank_us %>% slice_max(cases, n=1)`

The top 5 list: 
`r knitr::kable(rank_us %>% slice_max(cases, n=5) %>% select(-c(deaths,deaths_per_thou,rank_deaths,rank_deaths_mill)))`

US state with more cases/thousand: `r rank_us %>% slice_max(cases_per_thou, n=1) %>% select(-c(deaths,deaths_per_thou,rank_deaths,rank_deaths_mill))`

The top 5 cases/thousand list: 
`r knitr::kable(rank_us %>% slice_max(cases_per_thou, n=5) %>% select(-c(deaths,deaths_per_thou,rank_deaths,rank_deaths_mill)))`

The US State with more total deaths: `r rank_us %>% slice_max(deaths, n=1)`

The top 5 deaths list: 
`r knitr::kable(rank_us %>% slice_max(deaths, n=5) %>% select(-c(cases,cases_per_thou,rank,rank_mill)))`

US state with more deaths/thousand: `r rank_us %>% slice_max(deaths_per_thou, n=1)`

The top 5 deaths/thousand list: 
`r knitr::kable(rank_us %>% slice_max(deaths_per_thou, n=5) %>% select(-c(cases,cases_per_thou,rank,rank_mill)))`

-   The relation between deaths per thousand and cases per thousand don't seem obvious: 

```{r states, fig.height = 24, fig.width = 25}
a <- us_state_totals %>%
     ggplot() +
     scale_y_log10() +
     geom_col(aes(x = Province_State, y = cases_per_thou, fill=Province_State)) +coord_flip()

b <- us_state_totals %>%
     ggplot() +
     geom_col(aes(x = Province_State, y = deaths_per_thou, fill=Province_State)) +coord_flip()

grid.arrange(a,b)

```

##Data Modeling

-   To see if there is a relation between cases and deaths per thousand we'll use a linear model:

```{r linear}
modl <- lm(deaths_per_thou~cases_per_thou, data = us_state_totals)
summary(modl)
us_state_totals_pred <- us_state_totals %>% mutate(pred = predict(modl))
us_state_totals_pred %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```





## Conclussion


We've imported a dataset of Covid data from the beginning of the reported cases. By transforming the data and computing values we are able to makes plots that gives us a visual understanding of the evolution of the COVID Pandemic.
We've seen how the cases grew exponentially in the beginning and get flatter along the way. 
There is a statistically significant relationship between cases and deaths per thousand (p-value: 1.262e-05).

## Bias 


I have chosen a path that directs me on the direction of Us data as i leave in the United States instead of analizing the global dataset.