---
title: "NYPD shooting"
author: "Ernesto Medina"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

# Import libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("stringr") 
library("readr")
library(dplyr)
library(tidyr)
library(tidyverse)
library(lubridate)
library(gridExtra)
```


## NYPD shooting incident data

-   What neighborhood has more incidents?

-   Who is more affected?

-   Are shootings slowing down?




###Data source:

<https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>.

Get url for data:

```{r urls, message=TRUE}
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/"
file_name <- c("rows.csv?accessType=DOWNLOAD")
urls <- str_c(url_in, file_name)
urls
```

### Import data and print

```{r import_data, message=FALSE}
incidents <- read_csv(urls[1])
incidents
```

# Data Description

In this document we are presenting data from NYPD data source. We will go through a process of visual analysis and modeling. As we can see in the summary:

-   Data collected daily 2006-2020.

-   The data consists of 23585 rows in the dataset.

-   Column names are: `r colnames(incidents)`

-   STATISTICAL_MURDER_FLAG and INCIDENT_KEY will be used to differentciate between murders/non murder incidents (TRUE/FALSE flag) and individual/multiple victims (incident key).

```{r inspect, message=TRUE, echo=TRUE}
summary(incidents)

```


## Transform Data

-   Transform data to see relation between incidents and deaths



```{r by_date_and_combine, message=FALSE, echo=FALSE}
incidents <- incidents %>%
  select(-c(`INCIDENT_KEY`,`Latitude`,`Longitude`,`Lon_Lat`, `X_COORD_CD`, `Y_COORD_CD`,`JURISDICTION_CODE`,`LOCATION_DESC`))

incidents <- incidents %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE)) %>%
  unite("BORO_PRECINCT",
        c(BORO,PRECINCT),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE) %>%
  unite("VIC_AGE_SEX",
        c(VIC_AGE_GROUP,VIC_SEX),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
summary(incidents)
```

-   Columns we keep in dataset: `r colnames(incidents)`

## Analyzing data

```{r filter_data_and_group, message=FALSE, echo=FALSE}
deaths_by_boro <- incidents %>% 
  filter(STATISTICAL_MURDER_FLAG == TRUE) %>%
  group_by(BORO,STATISTICAL_MURDER_FLAG, OCCUR_DATE) %>%
  tally(name="murder_cases") %>%
  ungroup() 

assaults_by_boro <- incidents %>%
  filter(STATISTICAL_MURDER_FLAG == FALSE) %>%
  group_by(BORO,STATISTICAL_MURDER_FLAG, OCCUR_DATE) %>%
  tally(name="assault_cases") %>%
  ungroup() 

cases_by_boro <- incidents %>%
  group_by(BORO, OCCUR_DATE) %>%
  summarize(Total_cases= n(), murder_cases=sum(STATISTICAL_MURDER_FLAG, na.rm=TRUE)) %>%
  ungroup()

total_cases_by_boro <- incidents %>%
  group_by(BORO) %>%
  summarize(Total_cases= n(), murder_cases=sum(STATISTICAL_MURDER_FLAG, na.rm=TRUE)) %>%
  ungroup()

rates_by_boro <- cases_by_boro %>%
  arrange(OCCUR_DATE) %>%
  mutate(Murder_rate = murder_cases / Total_cases) %>%
  group_by(BORO) %>%
  summarize(cases= n(), murder_cases = sum(murder_cases), murder_rate=sum(murder_cases)/sum(Total_cases)) %>%
  ungroup()

rates_by_boro_date <- cases_by_boro %>%
  arrange(OCCUR_DATE) %>%
  mutate(Murder_rate = murder_cases / Total_cases)

cases_sex_group <- incidents %>%
  group_by(VIC_SEX) %>%
  summarize(Total_cases= n(), murder_cases=sum(STATISTICAL_MURDER_FLAG, na.rm=TRUE)) %>%
  ungroup()

cases_sex_group_by_date <- incidents %>%
  group_by(VIC_SEX, OCCUR_DATE) %>%
  summarize(Total_cases= n(), murder_cases=sum(STATISTICAL_MURDER_FLAG, na.rm=TRUE)) %>%
  ungroup()


```

### Boros deaths,cases and murder rate: `r knitr::kable(rates_by_boro)`


##


```{r boros, fig.height = 4, fig.width = 5, message=FALSE, echo=FALSE}
a <- total_cases_by_boro %>%
     ggplot() +
     scale_y_log10() +
     geom_col(aes(x = BORO, y = Total_cases, fill=BORO)) +coord_flip()

b <- total_cases_by_boro %>%
     ggplot() +
     geom_col(aes(x = BORO, y = murder_cases, fill=BORO)) +coord_flip()

grid.arrange(a,b)

```

## Visualizing and analyzing results


### Cases Murders comparison

-    Are murder and incidents related?


```{r cases_murder_comparison, message=FALSE, echo=FALSE}
cases_by_boro %>%
  ggplot(aes(x = OCCUR_DATE, y= Total_cases)) +
  xlab("Date") +
  geom_line(aes(color="Total_cases"))+
  geom_point(aes(color="Total_cases")) +
  geom_line(aes(y=murder_cases,color="murder_cases"))+
  geom_point(aes(y=murder_cases,color="murder_cases")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "Murder/cases by date for specific Boro in NYC", y=NULL)
```


##

## Data Modeling

-   To see if there is a relation between cases and deaths we'll use a linear model:

```{r linear_model, message=FALSE, echo=FALSE}
modl <- lm(murder_cases~Total_cases, data = cases_by_boro)
summary(modl)
```


##



### Model plot

```{r plot_model, message=FALSE, echo=FALSE}
cases_by_boro_pred <- cases_by_boro %>% mutate(pred = predict(modl))
cases_by_boro_pred %>%
  ggplot() +
  geom_point(aes(x = Total_cases, y = murder_cases), color = "blue") +
  geom_point(aes(x = Total_cases, y = pred), color = "red")
```



##


### Cases  comparison


```{r cases_by_boro, message=FALSE, echo=FALSE}
cases_by_boro %>%
  ggplot(aes(x = OCCUR_DATE, y= Total_cases, color=BORO)) +
  xlab("Date") +
  geom_line()+
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "Incident by date for all Boros in NYC", y=NULL)
```


### Murders comparison


```{r murders_by_boro, message=FALSE, echo=FALSE}
cases_by_boro %>%
  ggplot(aes(x = OCCUR_DATE, y= murder_cases, color=BORO)) +
  xlab("Date") +
  geom_line()+
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "Murder cases by date for all Boros in NYC", y=NULL)
```



### Murders Rate by victim's sex

```{r murder_rate, message=FALSE, echo=FALSE}
cases_sex_group_by_date %>%
  ggplot(aes(x = OCCUR_DATE, y= Total_cases, color=VIC_SEX)) +
  xlab("Date") +
  geom_line()+
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle=90)) +
  labs(title = "Murder rate by sex in NYC", y=NULL)
```





# Conclusion

Brooklyn is the Neighborhood with more incidents and murders in NYC with a murder rate of 0.195. The boro with highest murder rate is Staten Island with 0.205 although it can be moticated by a lower population. Male victims are the most affected in shootings in NYC by a factor of 10. Finally the incidents number has had an increase in the last year after some declines, probably directly related to covid-19.

As potential bias my inclination against violence and guns probably affects the analysis I've given in this report.
I will favor showing bad outputs that shows how bad is gun violence.
